Interrogating the linear variable importance of local explanations of black-box models with animated linear projections

### 2021-12-12 _paper.rmd
Manual and radial tour sections and grammerly edits. Emailing to Yang.

<!-- Introduce explanatory vs predictive modeling -->
Breiman and Shmueli [@breiman_statistical_2001; @shmueli_explain_2010] introduce the idea of distinguishing modeling based on its purpose; _explanatory_ modeling is done for some inferential purpose such as hypothesis testing, while _predictive_ modeling predicts new, out-of-sample, observations. This distinction draws attention to the divide between interpretable models and black-box models. In explanatory modeling, interpretability is key for drawing inferential conclusions. While predictive modeling may opt for potentially more accurate black-box models. The intended use has important implications for model selection and development.

@@

<!-- ## Data visualization tours -->
In multivariate data visualization, a _tour_ [@asimov_grand_1985; @buja_grand_1986; @lee_review_2021] is a sequence of linear projections of data onto a lower-dimensional space, typically 1-3D. Tours are viewed as an animation over small changes to the projection basis. Structure in a projection can then be explored visually to see which variables contribute to the formation of that structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the structural shape of the shadow change gleans insight into the shape and features of the object.

<!-- manual tours --> 
There are various types of tours, which are distinguished by the generation of projection bases. In a _manual_ tour [@cook_manual_1997; @spyrison_spinifex_2020] this path is defined by changing the contribution of a selected variable. <!--tours and models --> Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries [@wickham_visualizing_2015], and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis path [@lee_pptree_2013; @da_silva_projection_2021].

<!-- Purposed approach -->
The approach proposed below is to use the radial, manual tour to interrogate a local explanation. After the identifying an observation of interest, its explanation can be evaluated by testing the support of the structure identified by the explanation as the contributions of the variables are varied with the radial tour. We provide a free and open-source R package __cheem__ with an interactive application to facilitate analysis. We give case studies of toy and modern datasets for both classification and regression tasks.

@@

<!-- Paper structure -->
The remainder of this paper is organized as follows. The next section, [Local explanation statistics](#sec:explanations), covers the background of the local explanation, SHAP, and the traditional visuals produced from it. [Tours and the radial tour](#sec:tour) digs deeper into these animations of continuous linear projections. The section [Application Design](#sec:applicationdesign) discusses the layout of the application, how it facilitates analysis, preprocessing, and package infrastructure. The section [Case Studies](#sec:casestudies) illustrates several applications of this method. We conclude with a [Discussion](#sec:discussion) of the insights we draw from classification and regression tasks.

@@

<!-- Uses of local explanations -->
These instance-level explanations are used in a variety of ways depending on the data. In images, saliency maps overlay or offset a heatmap indicating which pixels were important [@simonyan_deep_2014]. For instance, the presence of snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words [@vanni_textual_2018]. In the case of numeric regression, they are used to explain variable additive contributions from model intercept to the observation's prediction [@ribeiro_why_2016].

@@

<!-- Illustration -->
Figure \@ref(fig:shapdistrbd) shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offensive and movement are more important for the offensive player, while defensive and power skills are more important to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions, such as goalkeepers or middle fielders. Panel c) shows a simplified breakdown plot [@gosiewska_ibreakdown_2019], where a local explanation is used to additively explain the difference from the intercept to the observations prediction. Such additive approaches will show an asymmetry in the variable ordering, so we opt to fix the order to that of panel b), namely, by decreasing the sum of the SHAP values.

@@

TODO:XXX
```{r}
##TODO:XXX
```

@@

<!-- Approximations of the spaces, PC1:2 -->
An approximation of these spaces is given as the first two principal components of their respective spaces. This is shown side-by-side next to model information, the prediction, and the observed response variable. The orientation and magnitude  of the variables are inscribed on a unit circle. While a single 2D projection will rarely encompass all of the structure of higher-dimensional spaces, it is a reasonable summarization real task at hand, the selection of observation, and nearby comparison.

@@

<!-- Segue to classification -->
Now that we have introduced the global view and corresponding cheem radial tour let's discuss the differences between the classification and regression cases.

@@

The regression case, we opt to color the global view on a statistic, the observation's residual, the log Mahanabis distance in data space (a measure of outlyingness), and the correlation of the attribution projection with the observed response. In the radial tour, the horizontal positions are the same, the  attribution projection, then varying as the radial tour changes basis. While the vertical position is fixed to the observed response variable and residuals for the middle and right panels, respectively. This changes the display from 1D density to a scatterplot. The basis is still one component, the horizontal position, independent with the vertical position.

@@

(ref:citeRf) @liaw_classification_2002
(ref:citeTs) @kominsarczyk_treeshap_2021
\begin{itemize}
	\item \textbf{Data:} a complete numerical matrix; explanatory and response variable. An optional categorical variable can be mapped to color and shape of obervations. Explanatory variables are scaled in visualization after modeling or creating local explanations. 
	\item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [(ref:citeRf)] for compatability with the local explanation which requires tree-based models.
	\item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP, that is applicable to tree-based models. This is done with the package \textbf{treeshap} [(ref:citeTs), hosted on GitHub only]. The global view shows all observations in attribution space, requiring the variable importance from \emph{all} observations rather than just one.
\end{itemize}

@@

In figure \@ref(fig:casepenguins), a misclassified point is contrasted with a correctly classified point of its observed class nearby in data-space. The attribution space from the tree SHAP local explanations is a more separable space, where the comparison is squarely in the middle of the orange distribution. The primary observation is in-between the predicted and observed clusters, a sign of uncertainty in the prediction. The tour varies the contribution of bill length (b_l) as this variable differs most from the contribution of the comparison observation. Downplaying the contribution of bill length is crucial to the linear explanation of this observation be misclassified.

@@

From figure \@ref(fig:casechocolates) we similarly see that attribution space is more separable relative to data-space. Interestingly there is not the class imbalance that we suspected; there are only 6 chocolates labeled as dark and predicted as milk, while 8 of the inverse case. Calories from fat is the variable with the largest difference in treeshap attribution between these points.

@@

The 2020 season FIFA [@leone_fifa_2020; @biecek_dalex_2018], contains many skill measurements of soccer/football players and wage information. After aggregation of the skill measurements, we regress the log wages [2020 euros] given just the skill aggregates. The model was fit from 5000 observations of the nine skill aggregates before being thinned to 500 players to mitigate occlusion and render time. We compare a leading offensive fielder (L. Messi) with that of a top defensive fielder (V. van Dijk), the same observations were used in figure \@ref(fig:shapdistrbd).

@@

With figure \@ref(fig:casefifa) we will test the premise of the local explanation. If we remove reaction and movement skills from the basis, then offense skills has almost singular importance for the explanation of the offensive player. We vary the contribution of offensive skills. In the tour (3rd frame of b), offensive skills moved, and Messi is no longer separated from the group. We also notice that accuracy has rotated into the frame, maintaining some separability.

@@

We have illustrated this method on random forest models using the tree SHAP local explanation, while it could be generally used with any compatible model-explanation pairing. We apply it to the classification and regression tasks. We have created an open-source __R__ package __cheem__, available on CRAN, to facilitate preprocessing and exploration with the described interactive application. Toy and real data are provided or upload your data after preprocessing.

@@

We would like to thank Professor Przemyslaw Biecek for his input early in the project and to the broader $\text{MI}^\text2$ lab group for the __DALEX__ ecosystem of __R__ and __Python__ packages. This research was supported by Australian government Research Training Program (RTP) scholarships.

>>>>

<!-- Introduce explanatory vs predictive modeling -->
Breiman and Shmueli [@breiman_statistical_2001; @shmueli_explain_2010] introduce the idea of distinguishing modeling based on its purpose; _explanatory_ modeling is done for some inferential purpose such as hypothesis testing, while _predictive_ modeling predicts new, out-of-sample, observations. This distinction draws attention to the divide between interpretable models and black-box models. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for potentially more accurate black-box models. The intended use has important implications for model selection and development.

@@

<!-- ## Data visualization tours -->
In multivariate data visualization, a _tour_ [@asimov_grand_1985; @buja_grand_1986; @lee_state_2021] is a sequence of linear projections of data onto a lower-dimensional space, typically 1-3D. Tours are viewed as an animation over small changes to the projection basis. Structure in a projection can then be explored visually to see which variables contribute to the formation of that structure. The intuition is similar to watching the shadow of a hidden 3D object change as the object is rotated; watching the structural shape of the shadow change gleans insight into the shape and features of the object.

<!-- manual tours --> 
There are various types of tours, which are distinguished by the generation of projection bases. In a _manual_ tour [@cook_manual_1997; @spyrison_spinifex_2020], the path is defined by changing the contribution of a selected variable. <!--tours and models --> Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries [@wickham_visualizing_2015], and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis path [@lee_pptree_2013; @da_silva_projection_2021].

<!-- Purposed approach -->
The proposed approach is to use the radial, manual tour to interrogate a local explanation. After identifying an observation of interest, its explanation can be evaluated by testing the support of the structure identified by the explanation as the contributions of the variables are varied with the radial tour. We provide a free and open-source R package __cheem__ with an interactive application to facilitate analysis. We give case studies of toy and modern datasets for both classification and regression tasks.

@@

<!-- Paper structure -->
The remainder of this paper is organized as follows. The following section, [Local explanation statistics](#sec:explanations), covers the background of the local explanation, SHAP, and the traditional visuals produced from it. [Tours and the radial tour](#sec:tour) digs deeper into these animations of continuous linear projections. The section [Application Design](#sec:applicationdesign) discusses the layout of the application, how it facilitates analysis, preprocessing, and package infrastructure. The section [Case Studies](#sec:casestudies) illustrates several applications of this method. We conclude with a [Discussion](#sec:discussion) of the insights we draw from classification and regression tasks.

@@

<!-- Uses of local explanations -->
These instance-level explanations are used in a variety of ways depending on the data. In images, saliency maps overlay or offset a heatmap indicating which pixels were important [@simonyan_deep_2014]. For instance, snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words [@vanni_textual_2018]. In the case of numeric regression, they are used to explain variable additive contributions from model intercept to the observation's prediction [@ribeiro_why_2016].

@@

<!-- Illustration -->
Figure \@ref(fig:shapdistrbd) shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offensive and movement are more important for the offensive player, while defensive and power skills are more informative to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions, such as goalkeepers or middle fielders. Panel c) shows a simplified breakdown plot [@gosiewska_ibreakdown_2019], where a local explanation is used to additively explain the difference from the intercept to the observations prediction. Such additive approaches will show an asymmetry in the variable ordering, so we opt to fix the order to panel b), namely, by decreasing the sum of the SHAP values.

@@

<!-- tours intro -->
A data visualization _tour_ animates many linear projections over small changes in the basis. One of the key features of the tour is the object permanence of the data points; one can track the relative changes of observations as the basis moves, as opposed to discretely jumping to an orthogonal view with no intermediate information. There  are various types of tours, which are distinguished by the selection of their basis paths [@lee_state_2021; @cook_grand_2008].

## Manual tours and the radial subset

<!-- Manual tour -->
The manual tour [@cook_manual_1997] defines its basis path by manipulating the basis contribution of a selected variable. A manipulation dimension is appended onto the projection plane, with a full contribution given to the selected variable. The bases are then selected based on rotating this newly created manipulation space. A crucial feature of the manual tour is that it allows users to control the variable contributions of the basis. This means that such manipulations can be selected and queued in advance or selected on the spot for human-in-the-loop analysis [@karwowski_international_2006]. However, this navigation is relatively time-consuming due to the huge volume of $p$-space. It is advisable to use this method to explore the sensitivity of the variable contribution to a previously identified feature of interest. In this case, the projection of the normalized explanations is the feature of interest.

<!-- radial tour -->
More generally, the manual tour can change the contribution of the variable in any of the display dimensions. We will with a specific more directed interaction, namely, a _radial_ tour. In a radial tour, the selected variable is allowed to change its magnitude of contribution but not its angle; it must move along the direction of its original contribution radius.

@@

<!-- Approximations of the spaces, PC1:2 -->
An approximation of these spaces is given as the first two principal components of their respective spaces. This is shown side-by-side next to model information, the prediction, and the observed response variable. The orientation and magnitude  of the variables are inscribed on a unit circle. While a single 2D projection will rarely encompass all of the structure of higher-dimensional spaces, it is a reasonable summarization, given the real task at hand, the selection of observations to explore further.

@@

<!-- Segue to classification -->
Now that we have introduced the global view and corresponding cheem radial tour, let's discuss the differences between the classification and regression cases.

@@

The regression case, we opt to color the global view on a statistic, the observation's residual, the log Mahanabis distance in data space (a measure of outlyingness), and the correlation of the attribution projection with the observed response. In the radial tour, the horizontal positions are the same, the  attribution projection, then varying as the radial tour changes basis. While the vertical position is fixed to the observed response variable and residuals for the middle and right panels, respectively. This changes the display from 1D density to a scatterplot. The basis is still one component, the horizontal position, independent of the vertical position.

@@

(ref:citeRf) @liaw_classification_2002
(ref:citeTs) @kominsarczyk_treeshap_2021
\begin{itemize}
	\item \textbf{Data:} a complete numerical matrix; explanatory and response variable. An optional categorical variable can be mapped to the color and shape of observations. Explanatory variables are scaled in visualization after modeling or creating local explanations. 
	\item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [(ref:citeRf)] for compatibility with the local explanation, which requires tree-based models.
	\item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP, that is applicable to tree-based models. This is done with the package \textbf{treeshap} [(ref:citeTs), hosted on GitHub only]. The global view shows all observations in attribution space, requiring the variable importance from \emph{all} observations rather than just one.
\end{itemize}

@@

In figure \@ref(fig:casepenguins), a misclassified point is contrasted with a correctly classified point of its observed class nearby in data-space. The attribution space from the tree SHAP local explanations is a more separable space, where the comparison is squarely in the middle of the orange distribution. The primary observation is in-between the predicted and observed clusters, a sign of uncertainty in the prediction. The tour varies the contribution of bill length (b_l) as this variable differs most from the contribution of the comparison observation. Downplaying the contribution of bill length is crucial to the linear explanation of this observation being misclassified.

@@

In figure \@ref(fig:casechocolates), we similarly see that attribution space is more separable relative to data-space. Interestingly there is not the class imbalance that we suspected; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. Calories from fat is the variable with the largest difference in treeshap attribution between these points.

@@

The 2020 season FIFA data [@leone_fifa_2020; @biecek_dalex_2018] contains many skill measurements of soccer/football players and wage information. After aggregation of the skill measurements, we regress the log wages [2020 euros] given just the skill aggregates. The model was fit from 5000 observations of the nine skill aggregates before being thinned to 500 players to mitigate occlusion and render time. We compare a leading offensive fielder (L. Messi) with that of a top defensive fielder (V. van Dijk), the same observations were used in figure \@ref(fig:shapdistrbd).

@@

With figure \@ref(fig:casefifa), we will test the premise of the local explanation. If we remove reaction and movement skills from the basis, then offense skills have almost singular importance for the explanation of the offensive player. We vary the contribution of offensive skills. In the tour (3rd frame of b), offensive skills moved, and Messi is no longer separated from the group. We also notice that accuracy has rotated into the frame, maintaining some separability.

@@

We have illustrated this method on random forest models using the tree SHAP local explanation, while it could be generally used with any compatible model-explanation pairing. We apply it to the classification and regression tasks. We have created an open-source __R__ package __cheem__, available on CRAN, to facilitate preprocessing and exploration with the described interactive application. Toy and real data are provided, or upload your data after preprocessing.

@@

We would like to thank Professor Przemyslaw Biecek for his input early in the project and to the broader $\text{MI}^\text2$ lab group for the __DALEX__ ecosystem of __R__ and __Python__ packages. This research was supported by Australian Government Research Training Program (RTP) scholarships.

### 2021-12-12 _paper.tex
Manual and radial tour sections and grammerly edits. Emailing to Yang.

Breiman and Shmueli (Breiman 2001; Shmueli 2010) introduce the idea of distinguishing modeling based on its purpose; \emph{explanatory} modeling is done for some inferential purpose such as hypothesis testing, while \emph{predictive} modeling predicts new, out-of-sample, observations. This distinction draws attention to the divide between interpretable models and black-box models. In explanatory modeling, interpretability is key for drawing inferential conclusions. While predictive modeling may opt for potentially more accurate black-box models. The intended use has important implications for model selection and development.

@@

There are various types of tours, which are distinguished by the generation of projection bases. In a \emph{manual} tour (Cook and Buja 1997; Spyrison and Cook 2020) this path is defined by changing the contribution of a selected variable. Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries (Wickham, Cook, and Hofmann 2015), and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis path (Y. D. Lee et al. 2013; Silva, Cook, and Lee 2021).

The approach proposed below is to use the radial, manual tour to interrogate a local explanation. After the identifying an observation of interest, its explanation can be evaluated by testing the support of the structure identified by the explanation as the contributions of the variables are varied with the radial tour. We provide a free and open-source R package \textbf{cheem} with an interactive application to facilitate analysis. We give case studies of toy and modern datasets for both classification and regression tasks.

@@

The remainder of this paper is organized as follows. The next section, \protect\hyperlink{sec:explanations}{Local explanation statistics}, covers the background of the local explanation, SHAP, and the traditional visuals produced from it. \protect\hyperlink{sec:tour}{Tours and the radial tour} digs deeper into these animations of continuous linear projections. The section \protect\hyperlink{sec:applicationdesign}{Application Design} discusses the layout of the application, how it facilitates analysis, preprocessing, and package infrastructure. The section \protect\hyperlink{sec:casestudies}{Case Studies} illustrates several applications of this method. We conclude with a \protect\hyperlink{sec:discussion}{Discussion} of the insights we draw from classification and regression tasks.

@@

These instance-level explanations are used in a variety of ways depending on the data. In images, saliency maps overlay or offset a heatmap indicating which pixels were important (Simonyan, Vedaldi, and Zisserman 2014). For instance, the presence of snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words (Vanni et al. 2018). In the case of numeric regression, they are used to explain variable additive contributions from model intercept to the observation's prediction (Ribeiro, Singh, and Guestrin 2016).

@@

Figure \ref{fig:shapdistrbd} shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offensive and movement are more important for the offensive player, while defensive and power skills are more important to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions, such as goalkeepers or middle fielders. Panel c) shows a simplified breakdown plot (Gosiewska and Biecek 2019), where a local explanation is used to additively explain the difference from the intercept to the observations prediction. Such additive approaches will show an asymmetry in the variable ordering, so we opt to fix the order to that of panel b), namely, by decreasing the sum of the SHAP values.

@@

TODO:XXX

@@

An approximation of these spaces is given as the first two principal components of their respective spaces. This is shown side-by-side next to model information, the prediction, and the observed response variable. The orientation and magnitude of the variables are inscribed on a unit circle. While a single 2D projection will rarely encompass all of the structure of higher-dimensional spaces, it is a reasonable summarization real task at hand, the selection of observation, and nearby comparison.

@@

Now that we have introduced the global view and corresponding cheem radial tour let's discuss the differences between the classification and regression cases.

@@

\begin{itemize}
    \item \textbf{Data:} a complete numerical matrix; explanatory and response variable. An optional categorical variable can be mapped to color and shape of obervations. Explanatory variables are scaled in visualization after modeling or creating local explanations. 
    \item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [Liaw and Wiener (2002)] for compatability with the local explanation which requires tree-based models.
    \item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP, that is applicable to tree-based models. This is done with the package \textbf{treeshap} [Kominsarczyk et al. (2021), hosted on GitHub only]. The global view shows all observations in attribution space, requiring the variable importance from \emph{all} observations rather than just one.
\end{itemize}

@@

In figure \ref{fig:casepenguins}, a misclassified point is contrasted with a correctly classified point of its observed class nearby in data-space. The attribution space from the tree SHAP local explanations is a more separable space, where the comparison is squarely in the middle of the orange distribution. The primary observation is in-between the predicted and observed clusters, a sign of uncertainty in the prediction. The tour varies the contribution of bill length (b\_l) as this variable differs most from the contribution of the comparison observation. Downplaying the contribution of bill length is crucial to the linear explanation of this observation be misclassified.

@@

From figure \ref{fig:casechocolates} we similarly see that attribution space is more separable relative to data-space. Interestingly there is not the class imbalance that we suspected; there are only 6 chocolates labeled as dark and predicted as milk, while 8 of the inverse case. Calories from fat is the variable with the largest difference in treeshap attribution between these points.

@@

The 2020 season FIFA (Leone 2020; Biecek 2018), contains many skill measurements of soccer/football players and wage information. After aggregation of the skill measurements, we regress the log wages {[}2020 euros{]} given just the skill aggregates. The model was fit from 5000 observations of the nine skill aggregates before being thinned to 500 players to mitigate occlusion and render time. We compare a leading offensive fielder (L. Messi) with that of a top defensive fielder (V. van Dijk), the same observations were used in figure \ref{fig:shapdistrbd}.

@@

With figure \ref{fig:casefifa} we will test the premise of the local explanation. If we remove reaction and movement skills from the basis, then offense skills has almost singular importance for the explanation of the offensive player. We vary the contribution of offensive skills. In the tour (3rd frame of b), offensive skills moved, and Messi is no longer separated from the group. We also notice that accuracy has rotated into the frame, maintaining some separability.

@@

We have illustrated this method on random forest models using the tree SHAP local explanation, while it could be generally used with any compatible model-explanation pairing. We apply it to the classification and regression tasks. We have created an open-source \textbf{R} package \textbf{cheem}, available on CRAN, to facilitate preprocessing and exploration with the described interactive application. Toy and real data are provided or upload your data after preprocessing.

@@

We would like to thank Professor Przemyslaw Biecek for his input early in the project and to the broader \(\text{MI}^\text2\) lab group for the \textbf{DALEX} ecosystem of \textbf{R} and \textbf{Python} packages. This research was supported by Australian government Research Training Program (RTP) scholarships.

@@

@@

@@

\leavevmode\hypertarget{ref-lee_review_2021}{}%
Lee, Stuart, Dianne Cook, Natalia Da Silva, Ursula Laa, Earo Wang, Nick Spyrison, and H. Sherry Zhang. 2021. {``A {Review} of the {State}-of-the-{Art} on {Tours} for {Dynamic} {Visualization} of {High}-Dimensional {Data}.''} \emph{arXiv Preprint arXiv:2104.08016}.

>>>>

Breiman and Shmueli (Breiman 2001; Shmueli 2010) introduce the idea of distinguishing modeling based on its purpose; \emph{explanatory} modeling is done for some inferential purpose such as hypothesis testing, while \emph{predictive} modeling predicts new, out-of-sample, observations. This distinction draws attention to the divide between interpretable models and black-box models. In explanatory modeling, interpretability is vital for drawing inferential conclusions. While predictive modeling may opt for potentially more accurate black-box models. The intended use has important implications for model selection and development.

@@

There are various types of tours, which are distinguished by the generation of projection bases. In a \emph{manual} tour (Cook and Buja 1997; Spyrison and Cook 2020), the path is defined by changing the contribution of a selected variable. Applying tours to models has been done in a couple of contexts. Specifically for exploring various statistical model fits and classification boundaries (Wickham, Cook, and Hofmann 2015), and using tree- and forest-based approaches as a projection pursuit index to generate a tour basis path (Y. D. Lee et al. 2013; Silva, Cook, and Lee 2021).

The proposed approach is to use the radial, manual tour to interrogate a local explanation. After identifying an observation of interest, its explanation can be evaluated by testing the support of the structure identified by the explanation as the contributions of the variables are varied with the radial tour. We provide a free and open-source R package \textbf{cheem} with an interactive application to facilitate analysis. We give case studies of toy and modern datasets for both classification and regression tasks.

@@

The remainder of this paper is organized as follows. The following section, \protect\hyperlink{sec:explanations}{Local explanation statistics}, covers the background of the local explanation, SHAP, and the traditional visuals produced from it. \protect\hyperlink{sec:tour}{Tours and the radial tour} digs deeper into these animations of continuous linear projections. The section \protect\hyperlink{sec:applicationdesign}{Application Design} discusses the layout of the application, how it facilitates analysis, preprocessing, and package infrastructure. The section \protect\hyperlink{sec:casestudies}{Case Studies} illustrates several applications of this method. We conclude with a \protect\hyperlink{sec:discussion}{Discussion} of the insights we draw from classification and regression tasks.

@@

These instance-level explanations are used in a variety of ways depending on the data. In images, saliency maps overlay or offset a heatmap indicating which pixels were important (Simonyan, Vedaldi, and Zisserman 2014). For instance, snow may be highlighted when distinguishing if a picture contains a wolf or husky. In text analysis, word-level contextual sentiment analysis can be used to highlight the sentiment and magnitude of influential words (Vanni et al. 2018). In the case of numeric regression, they are used to explain variable additive contributions from model intercept to the observation's prediction (Ribeiro, Singh, and Guestrin 2016).

@@

Figure \ref{fig:shapdistrbd} shows the SHAP values of these players. Panel a) shows these players receive a sizable difference in wages. Panel b) shows the underlying distribution of the SHAP attributions while permuting the explanatory variables, with the medians being the SHAP values. In the light of the player position, the difference in the variable importance makes sense; offensive and movement are more important for the offensive player, while defensive and power skills are more informative to the model for explaining the prediction of the defensive player. We would likewise expect the profile of variable importance to be unique for star players of other positions, such as goalkeepers or middle fielders. Panel c) shows a simplified breakdown plot (Gosiewska and Biecek 2019), where a local explanation is used to additively explain the difference from the intercept to the observations prediction. Such additive approaches will show an asymmetry in the variable ordering, so we opt to fix the order to panel b), namely, by decreasing the sum of the SHAP values.

@@

A data visualization \emph{tour} animates many linear projections over small changes in the basis. One of the key features of the tour is the object permanence of the data points; one can track the relative changes of observations as the basis moves, as opposed to discretely jumping to an orthogonal view with no intermediate information. There are various types of tours, which are distinguished by the selection of their basis paths (S. Lee et al. 2021; Cook et al. 2008).

\hypertarget{manual-tours-and-the-radial-subset}{%
\subsection{Manual tours and the radial subset}\label{manual-tours-and-the-radial-subset}}

The manual tour (Cook and Buja 1997) defines its basis path by manipulating the basis contribution of a selected variable. A manipulation dimension is appended onto the projection plane, with a full contribution given to the selected variable. The bases are then selected based on rotating this newly created manipulation space. A crucial feature of the manual tour is that it allows users to control the variable contributions of the basis. This means that such manipulations can be selected and queued in advance or selected on the spot for human-in-the-loop analysis (Karwowski 2006). However, this navigation is relatively time-consuming due to the huge volume of \(p\)-space. It is advisable to use this method to explore the sensitivity of the variable contribution to a previously identified feature of interest. In this case, the projection of the normalized explanations is the feature of interest.

More generally, the manual tour can change the contribution of the variable in any of the display dimensions. We will with a specific more directed interaction, namely, a \emph{radial} tour. In a radial tour, the selected variable is allowed to change its magnitude of contribution but not its angle; it must move along the direction of its original contribution radius.

@@

An approximation of these spaces is given as the first two principal components of their respective spaces. This is shown side-by-side next to model information, the prediction, and the observed response variable. The orientation and magnitude of the variables are inscribed on a unit circle. While a single 2D projection will rarely encompass all of the structure of higher-dimensional spaces, it is a reasonable summarization, given the real task at hand, the selection of observations to explore further.

@@

Now that we have introduced the global view and corresponding cheem radial tour, let's discuss the differences between the classification and regression cases.

@@

\begin{itemize}
    \item \textbf{Data:} a complete numerical matrix; explanatory and response variable. An optional categorical variable can be mapped to the color and shape of observations. Explanatory variables are scaled in visualization after modeling or creating local explanations. 
    \item \textbf{Model:} any model can be used with this method. Currently, we apply random forest models via the package \textbf{randomForest} [Liaw and Wiener (2002)] for compatibility with the local explanation, which requires tree-based models.
    \item \textbf{Local explanation:} any model-compatible linear explanation could be used. We apply tree SHAP, a more computationally efficient variant of SHAP, that is applicable to tree-based models. This is done with the package \textbf{treeshap} [Kominsarczyk et al. (2021), hosted on GitHub only]. The global view shows all observations in attribution space, requiring the variable importance from \emph{all} observations rather than just one.
\end{itemize}

@@

In figure \ref{fig:casepenguins}, a misclassified point is contrasted with a correctly classified point of its observed class nearby in data-space. The attribution space from the tree SHAP local explanations is a more separable space, where the comparison is squarely in the middle of the orange distribution. The primary observation is in-between the predicted and observed clusters, a sign of uncertainty in the prediction. The tour varies the contribution of bill length (b\_l) as this variable differs most from the contribution of the comparison observation. Downplaying the contribution of bill length is crucial to the linear explanation of this observation being misclassified.

@@

In figure \ref{fig:casechocolates}, we similarly see that attribution space is more separable relative to data-space. Interestingly there is not the class imbalance that we suspected; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. Calories from fat is the variable with the largest difference in treeshap attribution between these points.

@@

The 2020 season FIFA data (Leone 2020; Biecek 2018) contains many skill measurements of soccer/football players and wage information. After aggregation of the skill measurements, we regress the log wages {[}2020 euros{]} given just the skill aggregates. The model was fit from 5000 observations of the nine skill aggregates before being thinned to 500 players to mitigate occlusion and render time. We compare a leading offensive fielder (L. Messi) with that of a top defensive fielder (V. van Dijk), the same observations were used in figure \ref{fig:shapdistrbd}.

@@

With figure \ref{fig:casefifa}, we will test the premise of the local explanation. If we remove reaction and movement skills from the basis, then offense skills have almost singular importance for the explanation of the offensive player. We vary the contribution of offensive skills. In the tour (3rd frame of b), offensive skills moved, and Messi is no longer separated from the group. We also notice that accuracy has rotated into the frame, maintaining some separability.

@@

We have illustrated this method on random forest models using the tree SHAP local explanation, while it could be generally used with any compatible model-explanation pairing. We apply it to the classification and regression tasks. We have created an open-source \textbf{R} package \textbf{cheem}, available on CRAN, to facilitate preprocessing and exploration with the described interactive application. Toy and real data are provided, or upload your data after preprocessing.

@@

We would like to thank Professor Przemyslaw Biecek for his input early in the project and to the broader \(\text{MI}^\text2\) lab group for the \textbf{DALEX} ecosystem of \textbf{R} and \textbf{Python} packages. This research was supported by Australian Government Research Training Program (RTP) scholarships.

@@

\leavevmode\hypertarget{ref-cook_grand_2008}{}%
Cook, Dianne, Andreas Buja, Eun-Kyung Lee, and Hadley Wickham. 2008. {``Grand {Tours}, {Projection} {Pursuit} {Guided} {Tours}, and {Manual} {Controls}.''} In \emph{Handbook of {Data} {Visualization}}, 295--314. Berlin, Heidelberg: Springer Berlin Heidelberg. \url{https://doi.org/10.1007/978-3-540-33037-0_13}.

@@

\leavevmode\hypertarget{ref-karwowski_international_2006}{}%
Karwowski, Waldemar. 2006. \emph{International {Encyclopedia} of {Ergonomics} and {Human} {Factors}, -3 {Volume} {Set}}. CRC Press.

@@

\leavevmode\hypertarget{ref-lee_state_2021}{}%
Lee, Stuart, Dianne Cook, Natalia da Silva, Ursula Laa, Nicholas Spyrison, Earo Wang, and H. Sherry Zhang. 2021. {``The State-of-the-Art on Tours for Dynamic Visualization of High-Dimensional Data.''} \emph{WIREs Computational Statistics} n/a (n/a): e1573. \url{https://doi.org/10.1002/wics.1573}.
